\textbf{Proposal for bachelor thesis}\\ Exploration of the influence of graph parameters on the generalization error of Graph Neural Networks

### 2025-02-03 paper/chapters/background.tex
No code changes made.

\subsection{Supervised learning}
Machine learning can be roughly divided into three categories: supervised learning, unsupervised learning, and reinforcement learning~\cite{bishop2006pattern}.

@@

\subsection{Feed-forward neural networks}

@@

where $x$ is the input vector, $W^{(i)}$ and $b^{(i)}$ are the weights and biases of the $i$-th layer, and $f$ is the activation function. The output $y$ is the predicted output of the network. $L$ is denoted as the depth of the network.

\subsection{Loss function}
The loss function is a function that measures the difference between the predicted output of a model and the true output. The goal of training a machine learning model is to minimize the loss function, which is done by adjusting the parameters of the model using an optimization algorithm such as gradient descent. The choice of loss function depends on the task at hand. For classification tasks, the cross-entropy loss function is commonly used, while for regression tasks, the mean squared error loss function is often used.

@@

where $y$ is the true output, $\hat{y}$ is the predicted output, and $i$ is the index of the classes. The cross-entropy loss function is used to measure the difference between the true output and the predicted output of a classification model.

\subsection{Activation function}
An activation function is a non-linear function that is applied to the output of a neuron in a neural network. The activation function introduces non-linearity into the network, allowing it to learn complex patterns in the data. There are many activation functions that can be used in neural networks, such as the sigmoid function, the tanh function, and the ReLU function. The ReLU function is one of the most commonly used activation functions in deep learning because it is simple and computationally efficient. The ReLU function is defined as

@@

\subsection{Gradient Descent}
Gradient descent is an optimization algorithm used to minimize a function by iteratively moving in the direction of the steepest decrease of the function. The algorithm works by computing the gradient of the function at a given point and then updating the point in the direction of the negative gradient. The update rule is given by

@@

\subsection{Stochastic Gradient Descent}
Stochastic gradient descent (SGD) is a variant of gradient descent that uses a random sample of the training data to compute the gradient at each iteration. The algorithm works by randomly selecting a mini-batch of data points from the training dataset, computing the gradient of the loss function with respect to the mini-batch, and then updating the parameters based on the gradient. The update rule is given by

@@

\subsection{Early Stopping}
Early stopping is a technique used to prevent overfitting in machine learning models. The idea is to monitor the performance of the model on a validation dataset during training and stop training when the performance starts to degrade. This is done by keeping track of the validation loss and stopping training when the validation loss stops decreasing or starts to increase. Early stopping is used to prevent the model from memorizing the training data and to obtain a more generalizable model.

\subsection{Cross-validation}
Cross-validation is a technique used to evaluate the performance of a machine learning model. The idea is to split the dataset into multiple subsets, or folds, and train the model on one subset while evaluating it on the remaining subsets. This process is repeated multiple times, with each subset used as the test set once. The performance of the model is then averaged over all the folds to obtain an estimate of the model's generalization performance. Cross-validation is used to prevent overfitting and to obtain a more accurate estimate of the model's performance.

\subsection{Hyperparameter tuning}
Hyperparameter tuning is the process of selecting the optimal hyperparameters for a machine learning model. Hyperparameters are parameters that are set before the training process begins and control the behavior of the model. Examples of hyperparameters include the learning rate, batch size, number of hidden layers, and number of epochs. Hyperparameter tuning is an important step in the machine learning pipeline and can have a significant impact on the performance of the model. There are many techniques for hyperparameter tuning, such as grid search, random search, and Bayesian optimization. In this thesis, Bayesian optimization is used to tune the hyperparameters of the models.

@@

\subsection{Optimization algorithm}
An optimization algorithm is an algorithm used to minimize a function by iteratively updating the parameters of the model. The goal is to find the set of parameters that minimizes the loss function. Gradient descent is a common optimization algorithm used in machine learning, but there are many variants of gradient descent that are designed to improve convergence speed and stability. Some popular optimization algorithms include Adam~\cite{kingma2014adam}, RMSprop~\cite{graves2013generating}, and Adagrad~\cite{duchi2011adaptive}. These algorithms use different strategies to update the parameters of the model and can be more effective than standard gradient descent in certain situations.

\subsection{Generalization}
A key concept in evaluating model performance or generalization ability in such task is the empirical generalization error $E_{gen}$. This is a measure of how well the model generalizes to unseen data, which can be calculated as the difference between the training accuracy and the test accuracy. It is defined as

@@

\subsection{Graph Convolutional Networks (GCN)}

@@

\subsection{Simplified Graph Convolution (SGC)}

@@

\subsection{Graph Attention Networks (GAT)}
Graph Attention Networks (GAT) are a class of neural networks that use attention mechanisms to aggregate information from neighboring nodes. The GAT model updates the feature representation of a node by assigning different importance scores, known as attention coefficients, to its neighbors. The updated node feature $\mathbf{h}_i^{(l+1)}$ at layer $l+1$ is given by:

@@

\subsection{Graph Attention Networks v2 (GATv2)}
Graph Attention Networks v2 (GATv2)~\cite{brody2021attentive} is an improved version of the GAT model that introduces residual connections and multi-head attention. The GATv2 model is defined as

@@

\subsection{Message Passing Neural Networks (MPNN)}
Message Passing Neural Networks (MPNNs)~\cite{gilmer2017neural} have been the leading model for graph learning. The MPNN model operates by iteratively passing messages between nodes and updating their representations. At each iteration $t$, the message vector $m_i^{(t+1)}$ for node $i$ is computed as:

>>>>

\paragraph{Supervised learning}
Machine learning can be roughly divided into three categories: supervised learning, unsupervised learning, and reinforcement learning~\cite{bishop2006pattern}.

@@

\paragraph{Feed-forward neural networks}

@@

where $x$ is the input vector, $W^{(i)}$ and $b^{(i)}$ are the weights and biases of the $i$-th layer, and $f$ is the activation function. The output $y$ is the predicted output of the network. $L$ is denoted as the depth of the network. See Figure~\ref{fig:neural_network} for an illustration of a single neuron in a feed-forward neural network.

\usetikzlibrary{decorations.pathreplacing}
\begin{figure}[h!]
    \centering
\begin{tikzpicture}[>=stealth, scale=1.5, every node/.style={scale=1.2}]

    % Input signals
    \node at (-1.5, 1.5) (x1) {$x_1$};
    \node at (-1.5, 0.5) (x2) {$x_2$};
    \node at (-1.5, -1) (xm) {$x_m$};
    \node at (-1.5, -0.25) (dots) {$\vdots$};

    % Input brackets
    \draw[decorate, decoration={brace, amplitude=5pt}] (-1.8, -1.1) -- (-1.8, 1.6) node[midway, left, xshift=-0.2cm, scale=0.8] {Input signal};

    % Weights
    \node[draw, circle] at (0.3, 1.5) (w1) {$W_1$};
    \node[draw, circle] at (0.3, 0.5) (w2) {$W_2$};
    \node[draw, circle] at (0.3, -1) (wm) {$W_m$};

    \draw[->] (x1) -- (w1);
    \draw[->] (x2) -- (w2);
    \draw[->] (xm) -- (wm);

    % Sum node
    \node[draw, circle] at (2, 0) (sum) {$\Sigma$};

    % Weighted connections to sum
    \draw[->] (w1) -- (sum);
    \draw[->] (w2) -- (sum);
    \draw[->] (wm) -- (sum);

    % Bias
    \node[scale=0.8] at (2, 1.6) (biasText) {Bias};
    \node at (2, 1.2) (biasEq) {$b$};
    \draw[->] (biasEq) -- (sum);

    % Activation function
    \node[draw, rectangle] at (3.5, 0) (activation) {$\phi(\cdot)$};
    \node at (5, 0) (y) {$y$};
    \draw[->] (sum) -- (activation);
    \draw[->] (activation) -- (y);

    % Activation function label
    \node[scale=0.8] at (3.5, 0.6) {Activation function};

    % Weight label
    \node[scale=0.8] at (0.3, -1.7) {Weights};

    % Sum label
    \node[scale=0.8] at (2.05, -0.5) {Sum};
\end{tikzpicture}
\caption{An illustration of a single neuron in a feed-forward neural network. The neuron computes a weighted sum of the input signals, adds a bias term, and applies an activation function to produce the output.}
\label{fig:neural_network}
\end{figure}

\paragraph{Loss function}
The loss function is a function that measures the difference between the predicted output of a model and the true output. The goal of training a machine learning model is to minimize the loss function, which is done by adjusting the parameters of the model using an optimization algorithm such as gradient descent. The choice of loss function depends on the task at hand. For classification tasks, the cross-entropy loss function is commonly used, while for regression tasks, the mean squared error loss function is often used.

@@

where $y$ is the true output, $\hat{y}$ is the predicted output, and $i$ is the index of the classes. The cross-entropy loss function is used to measure the difference between the true output and the predicted output of a classification model. See Figure~\ref{fig:loss_function} for an example of a cost function surface in machine learning.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{loss_function.pdf}
    \caption{3D representation of a cost function surface in machine learning. Optimization algorithms aim to navigate this surface to reach the global minimum, improving model accuracy.}
    \label{fig:loss_function}
\end{figure}

\paragraph{Activation function}
An activation function is a non-linear function that is applied to the output of a neuron in a neural network. The activation function introduces non-linearity into the network, allowing it to learn complex patterns in the data. There are many activation functions that can be used in neural networks, such as the sigmoid function, the tanh function, and the ReLU function. The ReLU function is one of the most commonly used activation functions in deep learning because it is simple and computationally efficient. The ReLU function is defined as

@@

\paragraph{Gradient Descent}
Gradient descent is an optimization algorithm used to minimize a function by iteratively moving in the direction of the steepest decrease of the function. The algorithm works by computing the gradient of the function at a given point and then updating the point in the direction of the negative gradient. The update rule is given by

@@

\paragraph{Stochastic Gradient Descent}
Stochastic gradient descent (SGD) is a variant of gradient descent that uses a random sample of the training data to compute the gradient at each iteration. The algorithm works by randomly selecting a mini-batch of data points from the training dataset, computing the gradient of the loss function with respect to the mini-batch, and then updating the parameters based on the gradient. The update rule is given by

@@

\paragraph{Early Stopping}
Early stopping is a technique used to prevent overfitting in machine learning models. The idea is to monitor the performance of the model on a validation dataset during training and stop training when the performance starts to degrade. This is done by keeping track of the validation loss and stopping training when the validation loss stops decreasing or starts to increase. Early stopping is used to prevent the model from memorizing the training data and to obtain a more generalizable model.

\paragraph{Cross-validation}
Cross-validation is a technique used to evaluate the performance of a machine learning model. The idea is to split the dataset into multiple subsets, or folds, and train the model on one subset while evaluating it on the remaining subsets. This process is repeated multiple times, with each subset used as the test set once. The performance of the model is then averaged over all the folds to obtain an estimate of the model's generalization performance. Cross-validation is used to prevent overfitting and to obtain a more accurate estimate of the model's performance. See Figure~\ref{fig:cross-validation} for an example of 5-fold cross-validation.

\begin{center}
    \begin{tikzpicture}[scale=1.0]

    % Run 1
    \draw[thick] (0, 5) rectangle (5, 5.5);
    \fill[red!70] (0, 5) rectangle (1, 5.5);
    \draw[thick] (1, 5) -- (1, 5.5);
    \draw[thick] (2, 5) -- (2, 5.5);
    \draw[thick] (3, 5) -- (3, 5.5);
    \draw[thick] (4, 5) -- (4, 5.5);

    \node[right=6pt] at (5, 5.25) {run 1};

    % Run 2
    \draw[thick] (0, 4.5) rectangle (5, 5);
    \fill[red!70] (1, 4.5) rectangle (2, 5);
    \draw[thick] (1, 4.5) -- (1, 5);
    \draw[thick] (2, 4.5) -- (2, 5);
    \draw[thick] (3, 4.5) -- (3, 5);
    \draw[thick] (4, 4.5) -- (4, 5);

    \node[right=6pt] at (5, 4.75) {run 2};

    % Run 3
    \draw[thick] (0, 4) rectangle (5, 4.5);
    \fill[red!70] (2, 4) rectangle (3, 4.5);
    \draw[thick] (1, 4) -- (1, 4.5);
    \draw[thick] (2, 4) -- (2, 4.5);
    \draw[thick] (3, 4) -- (3, 4.5);
    \draw[thick] (4, 4) -- (4, 4.5);
    \node[right=6pt] at (5, 4.25) {run 3};

    % Run 4
    \draw[thick] (0, 3.5) rectangle (5, 4);
    \fill[red!70] (3, 3.5) rectangle (4, 4);
    \draw[thick] (1, 3.5) -- (1, 4);
    \draw[thick] (2, 3.5) -- (2, 4);
    \draw[thick] (3, 3.5) -- (3, 4);
    \draw[thick] (4, 3.5) -- (4, 4);
    \node[right=6pt] at (5, 3.75) {run 4};

    % Run 5
    \draw[thick] (0, 3) rectangle (5, 3.5);
    \fill[red!70] (4, 3) rectangle (5, 3.5);
    \draw[thick] (1, 3) -- (1, 3.5);
    \draw[thick] (2, 3) -- (2, 3.5);
    \draw[thick] (3, 3) -- (3, 3.5);
    \draw[thick] (4, 3) -- (4, 3.5);
    \node[right=6pt] at (5, 3.25) {run 5};

    \end{tikzpicture}
    \captionof{figure}{Example of 5-fold cross-validation, where the dataset is split into 5 folds and the model is trained and evaluated on each fold. The red portion of each fold represents the test set, while the white portion represents the training set. The performance of the model is averaged over all the folds to obtain an estimate of the model's generalization performance.}
    \label{fig:cross-validation}
\end{center}

\paragraph{Hyperparameter tuning}
Hyperparameter tuning is the process of selecting the optimal hyperparameters for a machine learning model. Hyperparameters are parameters that are set before the training process begins and control the behavior of the model. Examples of hyperparameters include the learning rate, batch size, number of hidden layers, and number of epochs. Hyperparameter tuning is an important step in the machine learning pipeline and can have a significant impact on the performance of the model. There are many techniques for hyperparameter tuning, such as grid search, random search, and Bayesian optimization. In this thesis, Bayesian optimization is used to tune the hyperparameters of the models.

@@

\paragraph{Optimization algorithm}
An optimization algorithm is an algorithm used to minimize a function by iteratively updating the parameters of the model. The goal is to find the set of parameters that minimizes the loss function. Gradient descent is a common optimization algorithm used in machine learning, but there are many variants of gradient descent that are designed to improve convergence speed and stability. Some popular optimization algorithms include Adam~\cite{kingma2014adam}, RMSprop~\cite{graves2013generating}, and Adagrad~\cite{duchi2011adaptive}. These algorithms use different strategies to update the parameters of the model and can be more effective than standard gradient descent in certain situations.

\paragraph{Generalization}
A key concept in evaluating model performance or generalization ability in such task is the empirical generalization error $E_{gen}$. This is a measure of how well the model generalizes to unseen data, which can be calculated as the difference between the training accuracy and the test accuracy. It is defined as

@@

\paragraph{Graph Convolutional Networks (GCN)}

@@

\paragraph{Simplified Graph Convolution (SGC)}

@@

\paragraph{Graph Attention Networks (GAT)}
Graph Attention Networks (GAT) are a class of neural networks that use attention mechanisms to aggregate information from neighboring nodes. The GAT model updates the feature representation of a node by assigning different importance scores, known as attention coefficients, to its neighbors. The updated node feature $\mathbf{h}_i^{(l+1)}$ at layer $l+1$ is given by:

@@

\paragraph{Graph Attention Networks v2 (GATv2)}
Graph Attention Networks v2 (GATv2)~\cite{brody2021attentive} is an improved version of the GAT model that introduces residual connections and multi-head attention. The GATv2 model is defined as

@@

\paragraph{Message Passing Neural Networks (MPNN)}
Message Passing Neural Networks (MPNNs)~\cite{gilmer2017neural} have been the leading model for graph learning. The MPNN model operates by iteratively passing messages between nodes and updating their representations. At each iteration $t$, the message vector $m_i^{(t+1)}$ for node $i$ is computed as:
